A Constraint Solving Approach to Parikh Images of Regular Languages

### 2024-02-29 introduction.tex
Clarify no backtrack
Nicer text

In applications involving regular languages, it is typically necessary
to consider the Parikh image not only of a single automaton, but of the
intersection of multiple automata. This problem arises in string solvers in
particular, as conjunctions of string constraints lead to the computation of
length images of products (intersections) of regular languages represented as
finite automata. The problem considered in this paper is thus:
{
    \centering
    \begin{tcolorbox}[colback=gray!5!white,colframe=gray!75!black,%
        title=Satisfiability of Parikh images of a product of automata,%
        width=0.8\linewidth]
        Given a finite set of deterministic, finite-state automata, $\Automaton_1, \ldots, \Automaton_n$
        over a finite, common, alphabet $\Set{a_1, \ldots, a_k}$,
        and a Presburger formula, $\phi(\VectorLiteral{x_1, \ldots, x_k})$, 
        decide if there is a word, $w$ accepted by the 
        product $\Automaton_1 \times \cdots \times \Automaton_n$, whose Parikh vector 
        $\VectorLiteral{m_1, \ldots, m_k}$ satisfies $\phi$.
    \end{tcolorbox}
  }

@@

Addressing these concerns, we have developed a calculus for Parikh images of
products of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image, and is generalised to an arbitrary homomorphism over
automata labels, including string lengths. This enables us to let both
calculations inform each other, eliminating unnecessary work, and pruning the
size of the partial products considered in the computation for a smaller memory
footprint. Moreover, the method can be used iteratively to tackle smaller chunks
of the product incrementally, thereby decreasing the memory footprint.

@@

        \begin{itemize}
            \item Enforce automata connectivity constraints lazily
            \item Use \emph{propagation} of constraints on the Parikh image to prune
            transitions from the automata that are incompatible with constraints
            \item Intelligently \emph{branch} on the presence\slash{}absence of key
             transitions to drive propagation of the connectivity constraint
             \item Compute products of automata lazily, after propagation
        \end{itemize}
    \end{tcolorbox}
  }

>>>>

In applications involving regular languages, it is typically necessary
to consider the Parikh image not only of a single automaton, but of the
intersection of multiple automata. This problem arises in string solvers in
particular, as conjunctions of string constraints lead to the computation of
length images of products (intersections) of regular languages represented as
finite automata. The problem considered in this paper is thus:

{
    \centering
    \begin{tcolorbox}[colback=gray!5!white,colframe=gray!75!black,%
        title=Satisfiability of Parikh images of a product of automata,%
        width=0.8\linewidth]
        Given a finite set of deterministic, finite-state automata, $\Automaton_1, \ldots, \Automaton_n$
        over a finite, common, alphabet $\Set{a_1, \ldots, a_k}$,
        and a Presburger formula, $\phi(\VectorLiteral{x_1, \ldots, x_k})$, 
        decide if there is a word, $w$ accepted by the 
        product $\Automaton_1 \times \cdots \times \Automaton_n$, whose Parikh vector 
        $\VectorLiteral{m_1, \ldots, m_k}$ satisfies $\phi$.
    \end{tcolorbox}
  }

@@

Addressing these concerns, we have developed a calculus for Parikh images of
products of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image, and is generalised to an arbitrary homomorphism over
automata labels, including string lengths. This enables us to let both
calculations inform each other, eliminating unnecessary work, and pruning the
size of the partial products considered in the computation for a smaller memory
footprint. Moreover, our approach allows us to do this precisely, without the expensive
fallbacks when approximations fail as in the case of~\cite{approximate-parikh}.

@@

        \begin{itemize}
            \item Enforce automata connectivity constraints lazily
            \item Use \emph{propagation} of constraints on the Parikh image to prune
            transitions from the automata that are incompatible with constraints
            \item Intelligently \emph{branch} on the presence\slash{}absence of key
             transitions to drive propagation of the connectivity constraint
             \item Compute products of automata lazily, after pruning transitions
             that violate constraints using propagation
        \end{itemize}
    \end{tcolorbox}
  }
